---
title: "Volcano plot"
author: "Jacqueline McCarty"
date: "2026-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(limma)
library(ggplot2)

# 1. Load the data
# We'll use the log-expression matrix and the final analysis metadata
expression_data <- read.csv("log_expression_matrix.csv", row.names = 1) # Samples as rows
metadata <- read.csv("final_analysis_dataset.csv")

# 2. Align Metadata and Expression Data
# limma expects genes as rows and samples as columns
expr_matrix <- t(expression_data)

# Ensure sample IDs in metadata match columns in expression matrix
# We assume 'sample_id' in metadata matches the row names of expression_data
metadata <- metadata %>% filter(sample_id %in% colnames(expr_matrix))
expr_matrix <- expr_matrix[, metadata$sample_id]

# 3. Define Groups
# Based on your dataset (GSE156063), we usually compare "SC2" (COVID) vs "no virus"
# We'll extract this from the 'char1' column or similar
metadata$group <- ifelse(grepl("SC2", metadata$char1), "COVID", "Control")
group <- factor(metadata$group, levels = c("Control", "COVID"))

# 4. Differential Expression Analysis with limma
design <- model.matrix(~group)
fit <- lmFit(expr_matrix, design)
fit <- eBayes(fit)

# Get results table
results <- topTable(fit, coef = 2, number = Inf) %>%
  rownames_to_column(var = "Gene")

# 5. Create the Volcano Plot
# We define thresholds for significance (e.g., p < 0.05 and |logFC| > 1)
results <- results %>%
  mutate(change = case_when(
    logFC > 1 & adj.P.Val < 0.05 ~ "Up-regulated",
    logFC < -1 & adj.P.Val < 0.05 ~ "Down-regulated",
    TRUE ~ "Not Significant"
  ))

ggplot(results, aes(x = logFC, y = -log10(adj.P.Val), color = change)) +
  geom_point(alpha = 0.4, size = 1.5) +
  scale_color_manual(values = c("Down-regulated" = "blue", 
                                "Not Significant" = "grey", 
                                "Up-regulated" = "red")) +
  theme_minimal() +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", color = "black") +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "black") +
  labs(title = "Volcano Plot: COVID-19 vs Control",
       subtitle = "Based on GSE156063 Cleaned Dataset",
       x = "Log2 Fold Change",
       y = "-Log10 Adjusted P-value") +
  theme(legend.position = "bottom")
```

X-axis (Log2 Fold Change): Points to the right (positive) are genes with higher expression in the COVID group; points to the left (negative) are higher in the Control group.

Y-axis (−log 10 P-value): The higher a point is on the plot, the more statistically significant the difference is.

Coloring:

Red: Significant genes up-regulated in COVID-19.

Blue: Significant genes down-regulated in COVID-19.

Grey: Genes that did not meet the statistical or fold-change threshold.



```{r}
library(ggplot2)
library(dplyr)
library(ggrepel)

# 1. Load the results and annotation data
results <- read.csv("gene dataset/Compliment dataset/DESeq2_COVID_vs_Healthy_results.csv")
# Load annotation to get Gene Symbols if available
annotation <- read.csv("gene dataset/Compliment dataset/counts_with_annotation.csv") %>%
  select(GeneID, Symbol) %>%
  distinct()

# 2. Merge results with gene symbols
df <- results %>%
  left_join(annotation, by = "GeneID") %>%
  filter(!is.na(padj)) # Remove rows with NA p-values

# 3. Define significance thresholds
lfc_threshold <- 1.0
pval_threshold <- 0.05

# 4. Categorize genes for coloring
df <- df %>%
  mutate(change = case_when(
    log2FoldChange > lfc_threshold & padj < pval_threshold ~ "Increased in COVID-19",
    log2FoldChange < -lfc_threshold & padj < pval_threshold ~ "Decreased in COVID-19",
    TRUE ~ "Not Significant"
  ))

# 5. Select top 10 genes to label (by significance)
top_genes <- df %>%
  filter(change != "Not Significant") %>%
  arrange(padj) %>%
  head(10)

# 6. Create the Volcano Plot
ggplot(df, aes(x = log2FoldChange, y = -log10(padj), color = change)) +
  geom_point(alpha = 0.4, size = 1.2) +
  scale_color_manual(values = c("Decreased in COVID-19" = "blue", 
                                "Not Significant" = "grey", 
                                "Increased in COVID-19" = "red")) +
  geom_vline(xintercept = c(-lfc_threshold, lfc_threshold), linetype = "dashed") +
  geom_hline(yintercept = -log10(pval_threshold), linetype = "dashed") +
  # Add labels for top genes
  geom_text_repel(data = top_genes, aes(label = Symbol), 
                  size = 3, color = "black", max.overlaps = 10) +
  theme_minimal() +
  labs(title = "Volcano Plot: COVID-19 vs Healthy",
       subtitle = "Significant genes labeled by Symbol",
       x = "Increase in Expression (Log2 Fold Change)",
       y = "Statistical Significance (-Log10 P-val)",
       color = "Expression Status",
       caption = "Source: National Center for Biotechnology Information") +
  theme(legend.position = "bottom")
```



```{r}
df <- read.csv("gene dataset/Compliment dataset/DESeq2_COVID_vs_Healthy_results.csv") %>%
  filter(!is.na(padj))

# 2. Use "Increased" and "Decreased" for labels
df <- df %>%
  mutate(Status = case_when(
    log2FoldChange > 1 & padj < 0.05 ~ "Increased in COVID-19",
    log2FoldChange < -1 & padj < 0.05 ~ "Decreased in COVID-19",
    TRUE ~ "No Significant Change"
  ))

# 3. Create the plot
ggplot(df, aes(x = log2FoldChange, y = -log10(padj), color = Status)) +
  geom_point(alpha = 0.4, size = 1.2) +
  scale_color_manual(values = c("Decreased in COVID-19" = "blue", 
                                "No Significant Change" = "grey", 
                                "Increased in COVID-19" = "red")) +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed", alpha = 0.5) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  theme_minimal() +
  labs(
    title = "Gene Expression Changes: COVID-19 vs Healthy Patients",
    x = "Increase in Expression (Log2 Fold Change)",
    y = "Statistical Significance (-Log10 P-val)",
    caption = "Source: National Center for Biotechnology Information"
  ) +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14),
    plot.caption = element_text(hjust = 0, face = "italic")
  )
```

```{r}
library(randomForest)
library(dplyr)
library(tidyr)

# 1. Load your data
counts <- read.csv("gene dataset/Compliment dataset/counts2_clean.csv", row.names = 1)
metadata <- read.csv("gene dataset/Compliment dataset/metadata2_clean.csv")
results <- read.csv("gene dataset/Compliment dataset/DESeq2_COVID_vs_Healthy_results.csv")

# 2. Transpose and fix Gene names
# We add "G_" to the start of every GeneID so R doesn't treat them as numbers
t_counts <- as.data.frame(t(counts))
colnames(t_counts) <- paste0("G_", colnames(t_counts))
t_counts$sample_id <- rownames(t_counts)

# Do the same for your top genes list
top_genes_list <- paste0("G_", results$GeneID[order(results$padj)])

# 3. Merge with metadata
data_for_rf <- merge(t_counts, metadata[, c("sample_id", "disease_state")], by = "sample_id")
data_for_rf$disease_state <- as.factor(data_for_rf$disease_state)

# 4. Corrected Function
run_rf_comparison <- function(n_genes, full_data, gene_ranking) {
  # Select the top N genes that actually exist in our data
  selected_genes <- intersect(gene_ranking, colnames(full_data))[1:n_genes]
  
  # Subset data to only include these genes and the target label
  rf_subset <- full_data[, c(selected_genes, "disease_state")]
  
  # Train model
  set.seed(123)
  # Using the data argument directly instead of a formula helps avoid naming errors
  rf_model <- randomForest(x = rf_subset[, -ncol(rf_subset)], 
                           y = rf_subset$disease_state, 
                           ntree = 500)
  
  cat("\n==============================================\n")
  cat("   Random Forest with Top", n_genes, "Genes\n")
  cat("==============================================\n")
  print(rf_model$confusion)
  
  # Calculate and print accuracy
  accuracy <- sum(diag(rf_model$confusion)) / sum(rf_model$confusion[,1:2])
  cat("\nOverall Accuracy:", round(accuracy * 100, 2), "%\n")
  
  return(rf_model)
}

# 5. Run the models
model_10  <- run_rf_comparison(10, data_for_rf, top_genes_list)
model_25  <- run_rf_comparison(25, data_for_rf, top_genes_list)
model_50  <- run_rf_comparison(50, data_for_rf, top_genes_list)
model_100 <- run_rf_comparison(100, data_for_rf, top_genes_list)
```

```{r}
library(randomForest)
library(caret)
library(dplyr)

# Load your datasets
counts <- read.csv("gene dataset/Compliment dataset/counts2_clean.csv", row.names = 1)
metadata <- read.csv("gene dataset/Compliment dataset/metadata2_clean.csv")
results <- read.csv("gene dataset/Compliment dataset/DESeq2_COVID_vs_Healthy_results.csv")

# Prepare Expression Data: Transpose and add "G_" prefix to Gene IDs
t_counts <- as.data.frame(t(counts))
colnames(t_counts) <- paste0("G_", colnames(t_counts))
t_counts$sample_id <- rownames(t_counts)

# Identify Top Genes sorted by significance (padj)
top_genes_list <- paste0("G_", results$GeneID[order(results$padj)])

# Merge with metadata to get 'disease_state'
data_for_rf <- merge(t_counts, metadata[, c("sample_id", "disease_state")], by = "sample_id")
data_for_rf$disease_state <- as.factor(data_for_rf$disease_state)
```


```{r}
# Subset data for top 10 genes
genes_10 <- intersect(top_genes_list, colnames(data_for_rf))[1:10]
df_10 <- data_for_rf[, c(genes_10, "disease_state")]

set.seed(123)
rf_10 <- randomForest(disease_state ~ ., data = df_10, ntree = 500, importance = TRUE)

# Print the Detailed Stats you requested
confusionMatrix(rf_10$predicted, df_10$disease_state)
```


```{r}
# Subset data for top 25 genes
genes_25 <- intersect(top_genes_list, colnames(data_for_rf))[1:25]
df_25 <- data_for_rf[, c(genes_25, "disease_state")]

set.seed(123)
rf_25 <- randomForest(disease_state ~ ., data = df_25, ntree = 500, importance = TRUE)

# Print the Detailed Stats
confusionMatrix(rf_25$predicted, df_25$disease_state)
```

```{r}
# Subset data for top 50 genes
genes_50 <- intersect(top_genes_list, colnames(data_for_rf))[1:50]
df_50 <- data_for_rf[, c(genes_50, "disease_state")]

set.seed(123)
rf_50 <- randomForest(disease_state ~ ., data = df_50, ntree = 500, importance = TRUE)

# Print the Detailed Stats
confusionMatrix(rf_50$predicted, df_50$disease_state)
```


```{r}
# Subset data for top 100 genes
genes_100 <- intersect(top_genes_list, colnames(data_for_rf))[1:100]
df_100 <- data_for_rf[, c(genes_100, "disease_state")]

set.seed(123)
rf_100 <- randomForest(disease_state ~ ., data = df_100, ntree = 500, importance = TRUE)

# Print the Detailed Stats
confusionMatrix(rf_100$predicted, df_100$disease_state)
```


```{r}
# Subset data for top 6 genes
genes_100 <- intersect(top_genes_list, colnames(data_for_rf))[1:6]
df_100 <- data_for_rf[, c(genes_100, "disease_state")]

set.seed(123)
rf_100 <- randomForest(disease_state ~ ., data = df_100, ntree = 500, importance = TRUE)

# Print the Detailed Stats
confusionMatrix(rf_100$predicted, df_100$disease_state)
```

```{r}
# Separate the features (x) and the target (y)
x_data <- data_for_rf[, colnames(data_for_rf) != "disease_state" & colnames(data_for_rf) != "sample_id"]
y_data <- data_for_rf$disease_state

set.seed(123)
# Passing x and y directly avoids the "Protection Stack" error
rf_all <- randomForest(x = x_data, y = y_data, ntree = 500)

print("--- Results for ALL GENES ---")
confusionMatrix(rf_all$predicted, y_data)

```

```{r}
#RANDOM 50 GENES ---
# This proves that the 'Top 50' were actually important
set.seed(456)
random_genes <- sample(colnames(t_counts), 50)
df_random <- data_for_rf[, c(random_genes, "disease_state")]

set.seed(123)
rf_random <- randomForest(disease_state ~ ., data = df_random, ntree = 500)
print("--- Results for RANDOM 50 GENES ---")
confusionMatrix(rf_random$predicted, df_random$disease_state)
```

