---
title: "EDA RNAseq - Jacqueline McCarty"
author: "Jacqueline McCarty"
date: "2026-04-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(pheatmap)
library(ggplot2)
library(stringr)

```


```{r}
counts <- read.csv("gene dataset/GSE156063_swab_gene_counts.csv", row.names = 1)
final_analysis_dataset <- read_csv("final_analysis_dataset.csv")

# check dimensions
dim(counts)
# summary
summary(as.vector(as.matrix(counts)))
```

We have ~16k genes across ~234 patients
Data is highly skewed (typical for RNA-seq)
```{r}
# log transform 
log_counts <- log1p(counts)

# plot distribution
hist(log_counts[,1], breaks = 50, main = "Gene Expression Distribution", col="skyblue")
```
	RNA data is skewed → log transform normalizes it
	
	
```{r}
gene_means <- rowMeans(log_counts)
gene_vars <- apply(log_counts, 1, var)

plot(gene_means, gene_vars,
     xlab = "Mean Expression",
     ylab = "Variance",
     main = "Mean vs Variance")
```
High variance genes = biologically important
	These will drive classification later
	
```{r}
# keep genes expressed in many samples
keep <- rowMeans(counts > 5) > 0.1
filtered_counts <- counts[keep, ]

log_filtered <- log1p(filtered_counts)

dim(filtered_counts)
```
	Removed noisy genes
	
	
	
```{r}
pca <- prcomp(t(log_filtered), scale. = TRUE)

# plot
plot(pca$x[,1], pca$x[,2],
     xlab = "PC1",
     ylab = "PC2",
     main = "PCA of Gene Expression")
```

	INTERPRET:
	•	If clusters appear → gene expression differs across patients
	•	Supports classification goal

```{r}
final_analysis_dataset$condition <- str_extract(
  final_analysis_dataset$`!Sample_characteristics_ch1`,
  "NEG|POS"
)
```

```{r}
pca_df <- data.frame(
  PC1 = pca$x[,1],
  PC2 = pca$x[,2],
  condition = final_analysis_dataset$condition   
)

# Plot
ggplot(pca_df, aes(x = PC1, y = PC2, color = condition)) +
  geom_point(size = 3, alpha = 0.8) +
  labs(title = "PCA of Gene Expression",
       color = "Condition") +
  theme_minimal()
```
```{r}
ggplot(pca_df, aes(x = PC1, y = PC2, color = final_analysis_dataset$sars_pcr_log10)) +
  geom_point(size = 3) +
  scale_color_viridis_c() +
  theme_minimal()
```

(idk what this means just yet)

- extracted disease labels from the metadata and used them to color PCA plots, allowing us to visually assess whether gene expression separates COVID and non-COVID samples




	
```{r}
# select top variable genes
top_genes <- order(gene_vars, decreasing = TRUE)[1:50]

pheatmap(log_filtered[top_genes, ],
         show_rownames = FALSE,
         main = "Top Variable Genes Heatmap")
```

	INTERPRET:
	•	Shows patterns across patients
	•	Helps identify immune-related variation
	

```{r}
sample_cor <- cor(log_filtered)

heatmap(sample_cor, main = "Sample Correlation")
```

INTERPRET:
	•	Similar samples cluster together
	•	Could indicate disease groups
	
```{r}
boxplot(log_filtered,
        outline = FALSE,
        main = "Expression Distribution per Sample")
```
INTERPRET:
	•	Checks normalization consistency
	
	
	
EDA shows there is meaningful variation in gene expression across patients
This suggests immune-related patterns may be detectable
This supports building a classifier independent of viral load
	
	
	
	EDA summary 
	The gene expression data shows a highly skewed distribution, which is typical for RNA-seq data, so a log transformation was applied to stabilize variance.

After filtering low-expression genes, we focused on the most informative features.

PCA revealed variation across samples, suggesting that gene expression profiles may separate different patient groups.

The heatmap of highly variable genes shows structured patterns across samples, indicating that certain genes may drive biological differences.

These findings support the feasibility of using gene expression data for classification tasks, which aligns with our goal of detecting immune response differences independent of viral load.



```{r}
final_analysis_dataset$covid_binary <- ifelse(
  final_analysis_dataset$condition == "POS", 1, 0
)
```

	
	
	
	
	immune_score vs viral load
	
```{r}
ggplot(final_analysis_dataset,
       aes(x = sars_pcr_log10, y = innate_score, color = condition)) +
  geom_point(size = 3, alpha = 0.7) +
  labs(
    title = "Immune Response vs Viral Load",
    x = "Viral Load (log10)",
    y = "Innate Immune Score"
  ) +
  theme_minimal()
```
If separation exists: Immune response differs across conditions even at similar viral loads


#Split into LOW vs HIGH viral load

```{r}
threshold <- median(final_analysis_dataset$sars_pcr_log10, na.rm = TRUE)

final_analysis_dataset$viral_group <- ifelse(
  final_analysis_dataset$sars_pcr_log10 < threshold,
  "Low",
  "High"
)
```

Boxplot 
```{r}
ggplot(final_analysis_dataset,
       aes(x = viral_group, y = innate_score, fill = condition)) +
  geom_boxplot() +
  labs(
    title = "Immune Response by Viral Load Group",
    x = "Viral Load Group",
    y = "Innate Immune Score"
  ) +
  theme_minimal()
```
“Even in low viral load samples, immune response varies across conditions”

#Logistic regression
Does immune_score matter after controlling for viral load?
```{r}
model <- glm(
  covid_binary ~ innate_score + sars_pcr_log10,
  data = final_analysis_dataset,
  family = "binomial"
)

summary(model)
```






only low viral load extra()

```{r}
low_data <- subset(final_analysis_dataset, viral_group == "Low")

model_low <- glm(
  covid_binary ~ innate_score,
  data = low_data,
  family = "binomial"
)

summary(model_low)
```





